---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
results <- readRDS("inst/benchmarks/results.rds")
```

# BackPropNN

<!-- badges: start -->

[![R-CMD-check](https://github.com/SinghRavin/BackPropNN/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/SinghRavin/BackPropNN/actions/workflows/R-CMD-check.yaml)

<!-- badges: end -->

The neural network algorithm trained by the process of backpropagation is implemented. The stochastic gradient descent (SGD) is adopted for this algorithm i.e., the weight and bias matrices are updated after learning from the errors of each data set point one by one. In this package, 3 layer (input, one hidden and output) neural network is considered. The mathematical equation involved are given below,

**Feed Forward:**

[H] = sigma([W_IH].[I] + [B_H])

[O] = sigma([W_HO].[H] + [B_O])

**Backpropagation:**

[delta_W\_HO] = (learning_rate)[Output_Errors]x[O(1-O)].[H_tranpose]

[delta_W\_IH] = (learning_rate)[Hiddden_Errors]x[H(1-H)].[I_tranpose]

[delta_B\_O] = (learning_rate)[Output_Errors]x[O(1-O)]

[delta_B\_H] = (learning_rate)[Hidden_Errors]x[H(1-H)]

Where, [] represents the matrix, x represents the Hadamard multiplication (element-wise), and . represents the usual matrix multiplication. H represents hidden matrix, O represents output matrix, W_IH represents weight matrix between input layer and hidden layer, W_HO represents weight matrix between hidden layer and output layer, B_H represents bias matrix for hidden layer, B_O represents bias matrix for output layer. The chosen activation function are Sigmoid or ReLU.

Weights are initialised randomly with variance 1/fan-in, and biases at zero. Both the R and the C++ engines draw from R's RNG stream, so `set.seed()` gives identical weights in either engine. The `epochs` argument controls how many complete passes are made over the data; it defaults to 1.

## Installation

You can install the development version of BackPropNN from GitHub like so:

``` r
# install.packages("devtools")
devtools::install_github("SinghRavin/BackPropNN")
```

## Example

This is a basic example which shows you how to solve a common problem:

# Simulated data - Logistics data.

```{r}
library(BackPropNN) # Loading the package.

set.seed(100)
num_obs <- 10000 # Number of observations.

# Setting coefficients values for the logit function.
beta0 <- -13.6
beta1 <- 0.10
beta2 <- 0.05

# Simulating the independent variables.
X1 <- runif(n=num_obs, min=18, max=60)
X2 <- runif(n=num_obs, min=100, max=250)
prob <- exp(beta0 + beta1*X1 + beta2*X2) / (1 + exp(beta0 + beta1*X1 + beta2*X2))

# Generating binary outcome variable.
Yvec <- rbinom(n=num_obs, size=1, prob=prob)

# Predictors are standardised. A sigmoid saturates when inputs are on the raw
# scale (X2 averages 175), which stalls learning in every engine compared here.
data <- data.frame(X1 = as.numeric(scale(X1)),
                   X2 = as.numeric(scale(X2)),
                   Y  = Yvec)
X  <- as.matrix(data[, 1:2])
Ym <- as.matrix(data[, 3])
```

# Running the functions of BackPropNN package.

```{r}
i <- 2 # number of input nodes
h <- 4 # number of hidden nodes
o <- 1 # number of output nodes
learning_rate <- 0.1 # The learning rate of the algorithm
activation_func <- "sigmoid" # the activation function
epochs <- 10 # number of complete passes over the data

set.seed(1)
nn_model <- back_propagation_training(i, h, o, learning_rate, activation_func,
                                      data, epochs = epochs)
```

# Summarizing the results of nn_model.

```{r}
plot(nn_model)
summary(nn_model)
```

## A note on what "fair" means here

`nnet` optimises with BFGS: one weight update per iteration, each using a
gradient computed over the whole data set. `BackPropNN` uses stochastic
gradient descent: one weight update per observation. Ten passes over
`r format(results$settings$num_obs, big.mark = ",")` rows is therefore
`r format(results$settings$num_obs * results$settings$E, big.mark = ",")`
updates for `BackPropNN` and `r results$settings$E` updates for `nnet`. The two
are not interchangeable, so the comparisons below are reported in two separate
ways:

- **Speed** is compared at an equal number of passes over the data, since a
  pass costs the same arithmetic in every engine. This isolates implementation
  efficiency, which is what the Rcpp rewrite was meant to address.
- **Accuracy** is compared with each engine run to convergence under its own
  optimiser.

All results below were produced by
[`inst/benchmarks/fair_benchmark.R`](inst/benchmarks/fair_benchmark.R) in a
single run on `r format(results$meta$date, "%d %B %Y")` (`r results$meta$os`,
`r results$meta$r`, nnet `r results$meta$versions[["nnet"]]`). They are read
from a saved file rather than re-run at knit time, so the published figures
stay fixed between builds. Re-run that script to refresh them.

# Speed: training, at an equal number of passes

```{r}
results$train
results$train_rel
```

# Speed: prediction

```{r}
results$pred
results$pred_rel
```

The R and C++ engines are the same algorithm, so given the same seed they must
produce the same weights. The benchmark script asserts this, and it held:
`r results$engines_agree`.

# Accuracy

Models are trained on 70% of the data and evaluated on the held-out 30%,
averaged over `r results$settings$n_rep` random initialisations. The C++ engine
is used for the repeated fits because it produces weights identical to the R
engine at a fraction of the cost.

```{r}
knitr::kable(results$accuracy, digits = 4)
```

# Cost against accuracy

The practitioner's question is what accuracy a model reaches and what it cost
to get there. This table pairs the two, and is directly comparable across
engines regardless of the SGD/BFGS difference. Training time is the median of
five fits at a fixed seed.

```{r}
knitr::kable(results$cost_accuracy, digits = 4, row.names = FALSE)
```

# What these results show

**The C++ rewrite is where the speed came from, and the gap to `nnet` is
modest.** At an equal number of passes over the data, the Rcpp engine is two
orders of magnitude faster than the pure R engine. It remains somewhat slower
than `nnet`, whose inner loop is compiled C that has been tuned for decades.
The interesting quantity is not that C++ beats interpreted R, which is
expected, but that a straightforward Armadillo implementation lands within a
small factor of a mature reference implementation. Timings come from a single
Windows machine and vary with system load, so the ordering is reliable but the
exact multiples are approximate.

**The bottleneck was the interpreted loop, not the arithmetic.** The pure R
engine allocates an S3 list on every activation call, resolves a closure on
every call, and coerces matrices once per row. The garbage collector runs
hundreds of times during a single benchmark; the C++ engine triggers it zero
times and allocates less memory than `nnet`. The algorithm is unchanged between
the two engines: same initialisation, same update order, same equations.

**Accuracy is competitive with the reference implementation.** With
standardised predictors, `BackPropNN` reaches a higher AUC and a lower MSE than
`nnet` under its own default settings on this problem, at a comparable training
cost.

**Per pass over the data, SGD extracts far more than BFGS.** At ten passes
`BackPropNN` has made
`r format(results$settings$num_obs * results$settings$E, big.mark = ",")`
weight updates and `nnet` has made `r results$settings$E`, which is why the
ten-pass rows differ so much. This is a property of the optimisers, not of the
implementations, and it is the reason speed and accuracy are reported under
separate protocols above.

# Known limitations

- The output layer uses the same activation as the hidden layer, so ReLU is not
  appropriate for binary classification with this implementation.
- The number of output nodes is nominally an argument but the prediction path
  assumes a single output.
- There is no convergence check, learning-rate schedule, or mini-batching; the
  training loop runs for exactly the number of epochs requested.
- Predictors are not standardised internally. As the comparison above shows,
  this matters a great deal for sigmoid activations, and is left to the user.
