---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# BackPropNN

<!-- badges: start -->

[![R-CMD-check](https://github.com/SinghRavin/BackPropNN/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/SinghRavin/BackPropNN/actions/workflows/R-CMD-check.yaml)

<!-- badges: end -->

The neural network algorithm trained by the process of backpropagation is implemented. The stochastic gradient descent (SGD) is adopted for this algorithm i.e., the weight and bias matrices are updated after learning from the errors of each data set point one by one. In this package, 3 layer (input, one hidden and output) neural network is considered. The mathematical equation involved are given below,

**Feed Forward:**

[H] = sigma([W_IH].[I] + [B_H])

[O] = sigma([W_HO].[H] + [B_O])

**Backpropagation:**

[delta_W\_HO] = (learning_rate)[Output_Errors]x[O(1-O)].[H_tranpose]

[delta_W\_IH] = (learning_rate)[Hiddden_Errors]x[H(1-H)].[I_tranpose]

[delta_B\_O] = (learning_rate)[Output_Errors]x[O(1-O)]

[delta_B\_H] = (learning_rate)[Hidden_Errors]x[H(1-H)]

Where, [] represents the matrix, x represents the Hadamard multiplication (element-wise), and . represents the usual matrix multiplication. H represents hidden matrix, O represents output matrix, W_IH represents weight matrix between input layer and hidden layer, W_HO represents weight matrix between hidden layer and output layer, B_H represents bias matrix for hidden layer, B_O represents bias matrix for output layer. The chosen activation function are Sigmoid or ReLU.

## Installation

You can install the development version of BackPropNN like so:

``` r
install.packages("BackPropNN")
```

## Example

This is a basic example which shows you how to solve a common problem:

# Simulated data - Logistics data.

```{r}
library(BackPropNN) # Loading the package.
num_obs <- 10000 # Number of observations.

# Setting coefficients values for the logit function.
beta0 <- -2.5
beta1 <- 0.02
beta2 <- 0.01

# Simulating the independent variables.
X1 <- runif(n=num_obs, min=18, max=60)
X2 <- runif(n=num_obs, min=100, max=250)
prob <- exp(beta0 + beta1*X1 + beta2*X2) / (1 + exp(beta0 + beta1*X1 + beta2*X2))

# Generating binary outcome variable.
Y <- rbinom(n=num_obs, size=1, prob=prob)

data <- data.frame(X1, X2, Y)
X <- as.matrix(data[1:ncol(data)-1])
Y <- as.matrix(data[,ncol(data)])
```

# Running the functions of BackPropNN package.

```{r}
set.seed(100)
i <- 2 # number of input nodes
h <- 4 # number of hidden nodes
o <- 1 # number of output nodes
learning_rate <- 0.1 # The learning rate of the algorithm
activation_func <- "sigmoid" # the activation function
nn_model <- back_propagation_training(i, h, o, learning_rate, activation_func, data)
nn_model_nnet <- nnet::nnet(X,Y,size=h, trace=FALSE)
```

# Summarizing the results of nn_model.

```{r}
plot(nn_model)
summary(nn_model)
print(nn_model) # This will print the benchmark comparison for training part.

# Running the benchmark comparison for prediction task.
bench::mark("BackPropNN"=feed_forward(data, nn_model),
            "R nnet" = as.numeric(stats::predict(nn_model_nnet,X, type="raw")),
                    relative = TRUE, check = FALSE)
```

We see that a slightly better AUC score is achieved by the R nnet package than BackPropNN, infact a better mse is achieved by R nnet (achieving slightly lower value than BackPropNN).

Also, BackPropNN (writing in R) consistently perform worse than R nnet package in terms of computational speed for both training and prediction tasks. 

# Now, let's check if the Rcpp version of BackPropNN package helps to improve the computational speed.

```{r}
# Running NN models using both versions.
nn_model_original <- back_propagation_training(i, h, o, learning_rate,
                                          activation_func, data)
nn_model_rcpp <- back_propagation_training_rcpp(i, h, o, learning_rate,
                                          activation_func, as.matrix(data))
nn_model_nnet <- nnet::nnet(X,Y,size=h, trace=FALSE)


# Running the benchmark comparison for training task.
bench::mark("Original"=back_propagation_training(i, h, o, learning_rate,
                                          activation_func, data),
            "R nnet"=nnet::nnet(X,Y,size=h, trace=FALSE),
            "Rcpp"=back_propagation_training_rcpp(i, h, o, learning_rate,
                                          activation_func, as.matrix(data)),
                    relative = TRUE, check = FALSE)

# Running the benchmark comparison for prediction task.
bench::mark("Original"=feed_forward(data, nn_model_original),
            "R nnet" = as.numeric(stats::predict(nn_model_nnet,X, type="raw")),
            "Rcpp"=feed_forward_rcpp(data, nn_model_rcpp),
                    relative = TRUE, check = FALSE)
```

We notice that for the training part, Rcpp version achieves best computational speed, however for the predicting part, R nnet package still takes the lead (performing just slightly better than Rcpp). The original package written in R lags behind in both training and prediction tasks.
