---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
results <- readRDS("inst/benchmarks/results.rds")
```

# BackPropNN

<!-- badges: start -->

[![R-CMD-check](https://github.com/SinghRavin/BackPropNN/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/SinghRavin/BackPropNN/actions/workflows/R-CMD-check.yaml)

<!-- badges: end -->

The neural network algorithm trained by the process of backpropagation is implemented. The stochastic gradient descent (SGD) is adopted for this algorithm i.e., the weight and bias matrices are updated after learning from the errors of each data set point one by one. In this package, 3 layer (input, one hidden and output) neural network is considered. The mathematical equation involved are given below,

**Feed Forward:**

[H] = sigma([W_IH].[I] + [B_H])

[O] = sigma([W_HO].[H] + [B_O])

**Backpropagation:**

[delta_W\_HO] = (learning_rate)[Output_Errors]x[O(1-O)].[H_tranpose]

[delta_W\_IH] = (learning_rate)[Hiddden_Errors]x[H(1-H)].[I_tranpose]

[delta_B\_O] = (learning_rate)[Output_Errors]x[O(1-O)]

[delta_B\_H] = (learning_rate)[Hidden_Errors]x[H(1-H)]

Where, [] represents the matrix, x represents the Hadamard multiplication (element-wise), and . represents the usual matrix multiplication. H represents hidden matrix, O represents output matrix, W_IH represents weight matrix between input layer and hidden layer, W_HO represents weight matrix between hidden layer and output layer, B_H represents bias matrix for hidden layer, B_O represents bias matrix for output layer. The chosen activation function are Sigmoid or ReLU.

Weights are initialised randomly with variance 1/fan-in, and biases at zero. Both the R and the C++ engines draw from R's RNG stream, so `set.seed()` gives identical weights in either engine. The `epochs` argument controls how many complete passes are made over the data; it defaults to 1.

## Installation

You can install the development version of BackPropNN from GitHub like so:

``` r
# install.packages("devtools")
devtools::install_github("SinghRavin/BackPropNN")
```

## Example

This is a basic example which shows you how to solve a common problem:

# Simulated data - Logistics data.

```{r}
library(BackPropNN) # Loading the package.

set.seed(100)
num_obs <- 10000 # Number of observations.

# Setting coefficients values for the logit function.
beta0 <- -13.6
beta1 <- 0.10
beta2 <- 0.05

# Simulating the independent variables.
X1 <- runif(n=num_obs, min=18, max=60)
X2 <- runif(n=num_obs, min=100, max=250)
prob <- exp(beta0 + beta1*X1 + beta2*X2) / (1 + exp(beta0 + beta1*X1 + beta2*X2))

# Generating binary outcome variable.
Yvec <- rbinom(n=num_obs, size=1, prob=prob)

# Predictors are standardised. A sigmoid saturates when inputs are on the raw
# scale (X2 averages 175), which stalls learning in every engine compared here.
data <- data.frame(X1 = as.numeric(scale(X1)),
                   X2 = as.numeric(scale(X2)),
                   Y  = Yvec)
```

# Running the functions of BackPropNN package.

```{r}
i <- 2 # number of input nodes
h <- 4 # number of hidden nodes
o <- 1 # number of output nodes
learning_rate <- 0.1 # The learning rate of the algorithm
activation_func <- "sigmoid" # the activation function
epochs <- 10 # number of complete passes over the data

set.seed(1)
nn_model <- back_propagation_training(i, h, o, learning_rate, activation_func,
                                      data, epochs = epochs)
```

# Summarizing the results of nn_model.

```{r}
nn_model
summary(nn_model)
```

## How the comparison is set up

`nnet` optimises with BFGS: one weight update per iteration, each using a
gradient over the whole data set. `BackPropNN` uses stochastic gradient
descent: one update per observation. An epoch and an iteration are therefore
not the same unit of work, and the two are compared under two protocols:

- **Speed** at an equal number of passes over the data. A pass costs the same
  arithmetic in every engine, so this isolates implementation efficiency.
- **Accuracy** with each engine run to convergence under its own optimiser.

Every number below comes from the same 70/30 split
(`r format(results$settings$n_train, big.mark = ",")` train,
`r format(results$settings$n_test, big.mark = ",")` test), the same seeds, and
the same high-resolution clock. Produced by
[`inst/benchmarks/fair_benchmark.R`](inst/benchmarks/fair_benchmark.R) in one
run on `r format(results$meta$date, "%d %B %Y")` (`r results$meta$os`,
`r results$meta$r`, nnet `r results$meta$versions[["nnet"]]`), and read from a
saved file so the published figures do not drift between builds.

# Speed: training, at an equal number of passes

```{r}
results$train
results$train_rel
```

# Speed: prediction

```{r}
results$pred
results$pred_rel
```

The R and C++ engines are the same algorithm, so given the same seed they must
produce the same weights. The benchmark script asserts this:
`r results$engines_agree`.

# Accuracy

Held-out AUC and MSE, averaged over `r results$settings$n_rep` random
initialisations. Both engines are run well past the point of improvement, so
convergence is demonstrated rather than assumed: `BackPropNN` is flat from 10
epochs onward, and `nnet` is identical at 500 and 2000 iterations because BFGS
reaches its tolerance and stops early.

```{r}
knitr::kable(results$accuracy, digits = 4)
```

# ROC curves on held-out data

Each engine at convergence: 100 epochs for `BackPropNN`, 500 BFGS iterations
for `nnet`. Single seed, so these AUCs sit close to but not exactly on the
averaged figures above.

```{r}
set.seed(100)
train_idx <- sample(seq_len(num_obs), size = round(0.7 * num_obs))
train <- data[train_idx, ]
test  <- data[-train_idx, ]

X_tr <- as.matrix(train[, 1:2]); Y_tr <- as.matrix(train[, 3])
X_te <- as.matrix(test[, 1:2])

set.seed(1)
fit_bp <- back_propagation_training_rcpp(i, h, o, learning_rate, activation_func,
                                         as.matrix(train), epochs = 100)
set.seed(1)
fit_nnet <- nnet::nnet(X_tr, Y_tr, size = h, maxit = 500, trace = FALSE)

invisible(pROC::roc(test$Y, as.numeric(feed_forward_rcpp(test, fit_bp)),
                    plot = TRUE, print.auc = TRUE, quiet = TRUE,
                    main = "ROC curve by BackPropNN (held-out)"))

invisible(pROC::roc(test$Y, as.numeric(stats::predict(fit_nnet, X_te, type = "raw")),
                    plot = TRUE, print.auc = TRUE, quiet = TRUE,
                    main = "ROC curve by R nnet (held-out)"))
```

# Cost against accuracy

What accuracy each engine reaches, and what it cost. Directly comparable
regardless of the SGD/BFGS difference. Training time is the median of five
seeded fits.

```{r}
knitr::kable(results$cost_accuracy, digits = 4, row.names = FALSE)
```

# What these results show

**The C++ rewrite removed the interpreter, not the arithmetic.** At an equal
number of passes the Rcpp engine is about 129 times faster than the pure R
engine, and about 2.4 times faster than `nnet`. Prediction is a tie with
`nnet`. The pure R engine allocates an S3 list on every activation call,
resolves a closure on every call, and coerces matrices once per row; its
garbage collector runs hundreds of times per benchmark, while the C++ engine
triggers it zero times and allocates less memory than `nnet`. The algorithm is
unchanged between the two engines: same initialisation, same update order, same
equations.

**Both engines converge, to different optima.** `BackPropNN` settles at an AUC
near 0.901 and does not move between 10 and 2000 epochs. `nnet` settles near
0.893, reached by 500 iterations and unchanged at 2000. The from-scratch
implementation converges to the better optimum, and reaches it roughly 28 times
faster.

**Most of that 28x is the optimiser, not the language.** Ten passes is 100,000
weight updates for SGD and ten for BFGS, so `BackPropNN` needs 50 times fewer
passes to converge. Multiply that by the 2.4x per-pass advantage and the gap
follows. This holds for a 17-parameter network on well-separated, linearly
generated data, which is where SGD is strongest. On a harder surface BFGS's
curvature information is exactly what pays off, and the ranking could reverse.

**Timings vary with system load.** They come from a single Windows machine, so
the ordering is reliable and the exact multiples are approximate.

# Known limitations

- The output layer uses the same activation as the hidden layer, so ReLU is not
  appropriate for binary classification with this implementation.
- The number of output nodes is nominally an argument but the prediction path
  assumes a single output.
- There is no convergence check, learning-rate schedule, or mini-batching; the
  training loop runs for exactly the number of epochs requested.
- Predictors are not standardised internally. As shown above, this matters a
  great deal for sigmoid activations, and is left to the user.
