---
title: "BackPropNN"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BackPropNN}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(BackPropNN)
```

This vignette shows how to use the package on three data sets: simulated
logistic data, a simulated AND gate, and the iris data. Timing and accuracy
comparisons against `nnet` are reported separately in the package README, which
is generated from a single benchmark run in
`inst/benchmarks/fair_benchmark.R`.

Two points apply throughout.

**Standardise the predictors.** The activations are sigmoid, which saturates
once inputs are far from zero. A predictor averaging 175 will drive the hidden
units to 0 or 1 on the first pass, at which point the derivative is
approximately zero and learning stalls. The package does not rescale inputs for
you.

**Set the number of epochs.** `epochs` is the number of complete passes over
the data and defaults to 1. One pass is rarely enough.

# Simulated data - Logistics data.

```{r}
set.seed(100)
num_obs <- 10000 # Number of observations

# Setting coefficients values for the logit function.
beta0 <- -13.6
beta1 <- 0.10
beta2 <- 0.05

# Simulating the independent variables.
X1 <- runif(n=num_obs, min=18, max=60)
X2 <- runif(n=num_obs, min=100, max=250)
prob <- exp(beta0 + beta1*X1 + beta2*X2) / (1 + exp(beta0 + beta1*X1 + beta2*X2))

# Simulating binary outcome variable.
Yvec <- rbinom(n=num_obs, size=1, prob=prob)

# Predictors are standardised before training.
data <- data.frame(X1 = as.numeric(scale(X1)),
                   X2 = as.numeric(scale(X2)),
                   Y  = Yvec)
```

# Running the functions of BackPropNN package.

```{r}
i <- 2 # number of input nodes
h <- 8 # number of hidden nodes
o <- 1 # number of output nodes
learning_rate <- 0.1 # The learning rate of the algorithm
activation_func <- "sigmoid" # the activation function

set.seed(1)
nn_model <- back_propagation_training(i, h, o, learning_rate, activation_func,
                                      data, epochs = 10)
```

# Summarizing the results of nn_model.

```{r}
plot(nn_model)
summary(nn_model)
```

The C++ engine is the same algorithm and returns the same weights when given
the same seed:

```{r}
set.seed(1)
nn_model_rcpp <- back_propagation_training_rcpp(i, h, o, learning_rate,
                                                activation_func,
                                                as.matrix(data), epochs = 10)

all.equal(unname(nn_model$weight_bias_matrices$weight_input_hidden),
          unname(nn_model_rcpp$weight_bias_matrices$weight_input_hidden))
```

# Simulated data - AND data.

The inputs here are already 0/1, so no rescaling is needed.

```{r}
set.seed(100)
num_obs <- 100000 # Number of observations
X1 <- sample(c(0,1),num_obs, replace = TRUE)
X2 <- sample(c(0,1),num_obs, replace = TRUE)
Y <- ifelse(X1==1 & X2==1, 1, 0)
data <- data.frame(X1,X2,Y)
```

# Running the functions of BackPropNN package.

```{r}
i <- 2 # number of input nodes
h <- 4 # number of hidden nodes
o <- 1 # number of output nodes
learning_rate <- 0.1 # The learning rate of the algorithm
activation_func <- "sigmoid" # the activation function

set.seed(1)
nn_model <- back_propagation_training(i, h, o, learning_rate, activation_func,
                                      data, epochs = 5)
```

# Summarizing the results of nn_model.

```{r}
plot(nn_model)
summary(nn_model)
```

# Real data - IRIS data.

A two-class problem, setosa against versicolor, using all four measurements as
predictors. The measurements are on comparable but non-trivial scales, so they
are standardised as well.

```{r}
data("iris") # Using iris data set.
iris2 <- iris[iris$Species != "virginica", ]
data <- data.frame(scale(iris2[, 1:4]),
                   Species = ifelse(iris2$Species == "setosa", 0, 1))
```

# Running the functions of BackPropNN package.

```{r}
i <- 4 # number of input nodes, i.e., number of X variables.
h <- 8 # number of hidden nodes
o <- 1 # number of output nodes
learning_rate <- 0.1 # The learning rate of the algorithm
activation_func <- "sigmoid" # the activation function

set.seed(1)
nn_model <- back_propagation_training(i, h, o, learning_rate, activation_func,
                                      data, epochs = 50)
```

Fewer than 100 rows here, so more passes are needed than in the simulated
examples above.

# Summarizing the results of nn_model.

```{r}
plot(nn_model)
summary(nn_model)
```

# Choosing the engine

`back_propagation_training()` and `back_propagation_training_rcpp()` implement
the same algorithm; the second is written in RcppArmadillo and is substantially
faster on larger data. The two differ only in how the data is passed:

```{r, eval = FALSE}
# R engine: takes a data frame
back_propagation_training(i, h, o, learning_rate, activation_func,
                          data, epochs = 10)

# C++ engine: takes a matrix
back_propagation_training_rcpp(i, h, o, learning_rate, activation_func,
                               as.matrix(data), epochs = 10)
```

Prediction follows the same pattern, with `feed_forward()` and
`feed_forward_rcpp()`. See the README for the timing comparison between the two
engines and against `nnet`.
